Vegetation regrowth/encroachment (gjengroing)

### 2023-09-11 gjengroing.Rmd

*Author and date:* Zander Venter

```{r}
Sys.Date()
```

@@

-   [NiN polygons](https://kartkatalog.miljodirektoratet.no/dataset/Details/2050) are used to identify reference areas with good ecological condition.
-   NIBIO's [AR5](https://kartkatalog.geonorge.no/metadata/arealressurskart-fkb-ar5-arealtyper/280bbd7a-5ce9-4c83-9e15-ac162cabd8a6) is used as the population sample for each ecosystem type. The population sample are the areas used to determine ecological condition relative to the reference areas. We also use the AR5 data to identify forest surrounding the population polygons for defining the lower limit of ecological condition index (ie. poor condition),
-   LiDAR-derived digital elevation model from Kartverket's [høydedata](https://hoydedata.no/LaserInnsyn/). This includes both a terrain (DTM) and surface mode (DSM). We calculate the canopy height model (DSM - DTM) to get the height of objects above the ground. From this we remove buildings, and what remains is vegetation - mostly trees but also some bushes or smaller woody plants.
-   Kartverket's [national 10m elevation model](https://kartkatalog.geonorge.no/metadata/dtm-10-terrengmodell-utm32/fd851873-f363-46f9-9fc6-bb1b403575df). We use this to stratify reference (good condition) and forest (poor condition) heights by elevational band. This is done in combination with bioclimatic zones - see next point.
-   Moen's [bioclimatic zones](https://artsdatabanken.no/Pages/181901/Bioklimatiske_soner). We use this to stratify reference (good condition) and forest (poor condition) heights by bioclimatic (also referred to as vegetation/climatic) zones.
-   [FKB building footprints](https://kartkatalog.geonorge.no/metadata/fkb-bygning/8b4304ea-4fb0-479c-a24d-fa225e2c6e97) are used to isolate vegetation in the LiDAR height data.
-   A [European satellite-based map](https://www.nature.com/articles/s41893-020-00609-y) of forest clear cuts is used for identifying AR5 forest patches in near-climax successional stages.
-   The [SSB 10km grid](https://kartkatalog.geonorge.no/metadata/statistisk-rutenett-5000m/32ac0653-d95c-446c-8558-bf9b79f4934e) is used for visualization purposes.
-   The regional delineation for Norway (five regions) are used for aggregating and reporting gjengroing condition values.

@@

### Aditional comments about the dataset

@@

![](images/gjengroing_schematic.jpg)

@@

The NiN polygons including "Naturlig åpne områder under skoggrensa", "Semi-naturlig mark", and "Våtmark" with good ecological condition are used to define a reference gjengroing state. We use the all-encompassing "Tilstand" variable assigned to each NiN polygon. The 50th percentile of LiDAR-derived vegetation heights within these polygons is used to define the upper limit (ie. 1) of the scaled indicator value. We cannot define local reference values based on proximity, because the NiN polygons are spatially biased and not close to all AR5 population polygons. Therefore we calculate regional reference values using elevation (300m bands) and bioclimatic zones as stratification. We calculate the mean reference value for each unique combination of elevation-bioclimatic zone. When calculating the index for each population polygon, the reference value is inhereted from the elecation-bioclimatic zone it falls within.

@@

We will visualize the elevation bands and bioclimatic zones that are used to define regional reference values for poor (forest vegetation height) and good (NiN vegetation height) ecological condition. For each unique spatial combination of these strata (n = 21), we calcualte a mean reference value later on in the analysis. Note that poor ecological condition is preferably calculated using a local reference (forest height within 200m of population polygon), but a regional reference is used where there are no nearby forest patches.

@@

Now we will aggregate up to the regional level for final reporting. We will use the same method as in the section above, except here we will use a different method to calculate the uncertainty value around the indicator values. The ea_spread function in eaTools uses a bootstrapping method to calculate the standard error in indicator values when aggregating from the polygon to regional level. Because we have hundreds of thousands of polygons per region, the resulting standard errors are extremely small. Therefore we will use the standard deviation in reference values per region. The standard deviation, as a percentage of the average reference value is used to calcualte the uncertainty in the scaled indicator value.

@@

-   **Selection of the reference and population areas.** Here we were limited to using NiN and AR5 because these are arguably the best datasets currently available for this exercise. However, both NiN and AR5 are spatially and thematically biased in some ways. Therefore, it may be worth exploring alternative datasets in the future. NiN polygons are biased towards low lying areas which are generally warmer and more productive. For a given bioclimatic-elevation strata, this could lead to higher reference values for vegetation height than one would expect with a representative reference sample. This could consequently result in high elevation population polygons appearing to have very good condition ebcause they are being compared with (relatively) high productivity polygons in the same bioclimatic-elevation strata. This is a possible explanation for why the high elevation parts of norway appear to be in very good ecological condition.

@@

-   **Quantifying regional vs local references for poor and good condition.** For good condition, we used NiN polygons and were restricted to these due to lack of better data. However, for poor condition we use climax successional stage for trees (as explained above) within 200m of each population polygon. For polygons with no surrounding forest, we impute with regional reference values (mean of forest height in elevation-bioclamic zone). The choice of 200m buffer zone was largely based on computational limitations. Calculating the average LiDAR vegetation height within 200m of nearly 2 million complex plygons takes time, even in GEE. It may make more ecological sense to define a larger buffer zone to use for defining reference for poor condition. However, longer processing times need to be expected.

@@

After exporting the final polygon-level indicator scores, we imported to QGIS and explored how they look at the landscape scale. We chose random locations spread across the 5 regions in Norway. The figures below show våtmark and åpne (semi- and naturlig-åpne) ecosysystem polygons from AR5. The polygons have jaggered edges because we chose a low resolution geometry export from GEE to save computing time and minimize storage space. Each polygon is colored based on the final scaled gjengroing indicator score (0 to 1). The polygons are also labelled with the median LiDAR-derived forest height. An orthophoto from Norgeibilder is displayed in the background for reference.

In the first example from midt-norge, we see a våtmark polygon colored red (ie. in poor condition) due to the growth of a forest in the northern section. It looks like a plantation forest and possibly occured after the wetland was drained. The height of the forest vegetation equates to a median of 2.4m which is higher than the reference value for this region (1.09m) and almost as high as the mature forest height reference of 2.8m. Therefore it gets a scaped indicator score of 0.1. In contrast the wetland plygon in the middle of the image is in good condition because it has a median vegetation height of 1.1m which is only fractionally higher than the regional reference for good ecological condition.

![](images/val_1.jpg)

The second example from Sørlandet shows three AR5 semi-naturlig åpne polygons with different indicator values. The lower polygon has poor ecological condition because its median vegetation height is 6.2m which is close to the local reference value for mature forest (7.9m). Although the centeral polygon appears to have greater coverage of forest in the orthophoto, the LiDAR data tells us that it is mostly covered by short trees and results in a median vegetation height of 2.8 - therefore good condition.

![](images/val_2.jpg)

@@

![](images/val_3.jpg)

@@

![](images/val_4.jpg)

The fifth and final example from further north in Nord-norge shows two våtmark polygons with contrasting condition scores, yet exactly the same median vegetation heights of 0.6m. The difference in the scaled iindicator value is due to different local reference values for surrounding forest. The polygon on the left has surrounding forest (within 200m buffer zone) with median height 0.5m, while the polygon on the right has surrounding forest of 0.7m. This creates a contrast in the final indicator valye and illustrates the effect that a local reference value can have on defining good and poor ecological condition.

![](images/val_5.jpg)

>>>>

*Author and date:*

Zander Venter

September 2023

@@

-   [NiN polygons](https://kartkatalog.miljodirektoratet.no/dataset/Details/2050) are used to identify reference areas with good ecological condition.
-   NIBIO's [AR5](https://kartkatalog.geonorge.no/metadata/arealressurskart-fkb-ar5-arealtyper/280bbd7a-5ce9-4c83-9e15-ac162cabd8a6) is used as the population sample for each ecosystem type. The population sample are the areas used to determine ecological condition relative to the reference areas. We also use the AR5 data to identify forest surrounding the population polygons for defining the lower limit of ecological condition index (i.e.. poor condition),
-   LiDAR-derived digital elevation model from Kartverket's [høydedata](https://hoydedata.no/LaserInnsyn/). This includes both a terrain (DTM) and surface mode (DSM). We calculate the canopy height model (DSM - DTM) to get the height of objects above the ground. From this we remove buildings, and what remains is vegetation - mostly trees but also some bushes or smaller woody plants.
-   Kartverket's [national 10m elevation model](https://kartkatalog.geonorge.no/metadata/dtm-10-terrengmodell-utm32/fd851873-f363-46f9-9fc6-bb1b403575df). We use this to stratify reference (good condition) and forest (poor condition) heights by elevational band. This is done in combination with bioclimatic zones - see next point.
-   Moen's [bioclimatic zones](https://artsdatabanken.no/Pages/181901/Bioklimatiske_soner). We use this to stratify reference (good condition) and forest (poor condition) heights by bioclimatic (also referred to as vegetation/climatic) zones.
-   [FKB building footprints](https://kartkatalog.geonorge.no/metadata/fkb-bygning/8b4304ea-4fb0-479c-a24d-fa225e2c6e97) are used to isolate vegetation in the LiDAR height data.
-   A [European satellite-based map](https://www.nature.com/articles/s41893-020-00609-y) of forest clear cuts is used for identifying AR5 forest patches in near-climax successional stages.
-   The [SSB 10km grid](https://kartkatalog.geonorge.no/metadata/statistisk-rutenett-5000m/32ac0653-d95c-446c-8558-bf9b79f4934e) is used for visualization purposes.
-   The regional delineation for Norway (five regions) are used for aggregating and reporting gjengroing condition values.

@@

### Additional comments about the dataset

@@

![](images/gjengroing_schematic.jpg){width="70%"}

@@

The NiN polygons including "Naturlig åpne områder under skoggrensa", "Semi-naturlig mark", and "Våtmark" with good ecological condition are used to define a reference gjengroing state. We use the all-encompassing "Tilstand" variable assigned to each NiN polygon. The 50th percentile of LiDAR-derived vegetation heights within these polygons is used to define the upper limit (ie. 1) of the scaled indicator value. We cannot define local reference values based on proximity, because the NiN polygons are spatially biased and not close to all AR5 population polygons. Therefore we calculate regional reference values using elevation (300m bands) and bioclimatic zones as stratification. We calculate the mean reference value for each unique combination of elevation-bioclimatic zone. When calculating the index for each population polygon, the reference value is inherited from the elevation-bioclimatic zone it falls within.

@@

We will visualize the elevation bands and bioclimatic zones that are used to define regional reference values for poor (forest vegetation height) and good (NiN vegetation height) ecological condition. For each unique spatial combination of these strata (n = 21), we calculate a mean reference value later on in the analysis. Note that poor ecological condition is preferably calculated using a local reference (forest height within 200m of population polygon), but a regional reference is used where there are no nearby forest patches.

@@

Now we will aggregate up to the regional level for final reporting. We will use the same method as in the section above, except here we will use a different method to calculate the uncertainty value around the indicator values. The ea_spread function in eaTools uses a bootstrapping method to calculate the standard error in indicator values when aggregating from the polygon to regional level. Because we have hundreds of thousands of polygons per region, the resulting standard errors are extremely small. Therefore we will use the standard deviation in reference values per region. The standard deviation, as a percentage of the average reference value is used to calculate the uncertainty in the scaled indicator value.

@@

-   **Selection of the reference and population areas.** Here we were limited to using NiN and AR5 because these are arguably the best datasets currently available for this exercise. However, both NiN and AR5 are spatially and thematically biased in some ways. Therefore, it may be worth exploring alternative datasets in the future. NiN polygons are biased towards low lying areas which are generally warmer and more productive. For a given bioclimatic-elevation strata, this could lead to higher reference values for vegetation height than one would expect with a representative reference sample. This could consequently result in high elevation population polygons appearing to have very good condition because they are being compared with (relatively) high productivity polygons in the same bioclimatic-elevation strata. This is a possible explanation for why the high elevation parts of Norway appear to be in very good ecological condition.

@@

-   **Quantifying regional vs local references for poor and good condition.** For good condition, we used NiN polygons and were restricted to these due to lack of better data. However, for poor condition we use climax successional stage for trees (as explained above) within 200m of each population polygon. For polygons with no surrounding forest, we impute with regional reference values (mean of forest height in elevation-bioclamic zone). The choice of 200m buffer zone was largely based on computational limitations. Calculating the average LiDAR vegetation height within 200m of nearly 2 million complex polygons takes time, even in GEE. It may make more ecological sense to define a larger buffer zone to use for defining reference for poor condition. However, longer processing times need to be expected.

@@

After exporting the final polygon-level indicator scores, we imported to QGIS and explored how they look at the landscape scale. We chose random locations spread across the 5 regions in Norway. The figures below show våtmark and åpne (semi- and naturlig-åpne) ecosystem polygons from AR5. The polygons have jagged edges because we chose a low resolution geometry export from GEE to save computing time and minimize storage space. Each polygon is colored based on the final scaled gjengroing indicator score (0 to 1). The polygons are also labelled with the median LiDAR-derived forest height. An orthophoto from Norgeibilder is displayed in the background for reference.

In the first example from midt-norge, we see a våtmark polygon colored red (ie. in poor condition) due to the growth of a forest in the northern section. It looks like a plantation forest and possibly occurred after the wetland was drained. The height of the forest vegetation equates to a median of 2.4m which is higher than the reference value for this region (1.09m) and almost as high as the mature forest height reference of 2.8m. Therefore it gets a scaped indicator score of 0.1. In contrast the wetland polygon in the middle of the image is in good condition because it has a median vegetation height of 1.1m which is only fractionally higher than the regional reference for good ecological condition.

![](images/val_1.jpg){width="50%"}

The second example from Sørlandet shows three AR5 semi-naturlig åpne polygons with different indicator values. The lower polygon has poor ecological condition because its median vegetation height is 6.2m which is close to the local reference value for mature forest (7.9m). Although the central polygon appears to have greater coverage of forest in the orthophoto, the LiDAR data tells us that it is mostly covered by short trees and results in a median vegetation height of 2.8 - therefore good condition.

![](images/val_2.jpg){width="50%"}

@@

![](images/val_3.jpg){width="50%"}

@@

![](images/val_4.jpg){width="50%"}

The fifth and final example from further north in Nord-norge shows two våtmark polygons with contrasting condition scores, yet exactly the same median vegetation heights of 0.6m. The difference in the scaled indicator value is due to different local reference values for surrounding forest. The polygon on the left has surrounding forest (within 200m buffer zone) with median height 0.5m, while the polygon on the right has surrounding forest of 0.7m. This creates a contrast in the final indicator value and illustrates the effect that a local reference value can have on defining good and poor ecological condition.

![](images/val_5.jpg){width="50%"}
